---
title: "Adult ADHD Review Results"
author: "Joey W. Trampush, Ph.D."
format: 
  typst: 
    toc: false
  html: 
    toc: true
  docx:
    toc: true
execute: 
  echo: false # Don't show code
  warning: false # Don't show warnings
  message: false # Don't show messages
  include: true # DO show results
---

```{r setup}
# Load necessary libraries
library(gt)
library(tibble)
library(dplyr)
library(readr)
library(vroom)

# Source the custom table function
source("R/create_demographics_table.R")
source("R/create_table.R")
```

## Rating Scale and Diagnostic Interviews

```{r warning=FALSE}
# For the first table, we'll use knitr::kable instead of gt for better cross-format compatibility
adhd_diagnostic_tools <- read_csv("outputs/tables/adhd-diagnostic-tools.csv")

# Select only key columns for better readability across formats
df_display <- adhd_diagnostic_tools %>%
  select(Study, Measure_Name, Sensitivity, Specificity, Diagnostic_Accuracy, Notes)

# Create the table with knitr::kable
knitr::kable(
  df_display,
  col.names = c("Study", "Diagnostic Measure", "Sensitivity", "Specificity", "Diagnostic Accuracy", "Notes"),
  caption = "Rating Scales and Diagnostic Interviews for Adult ADHD"
)
```

## Neuroimaging

```{r}
# Create a data frame using tibble::tribble for clarity
demographics_data_neuroimaging <- tibble::tribble(
  ~Study, ~Sample_Size, ~Age, ~Gender, ~ADHD_Subtypes, ~Imaging_Modality, ~Notes,
  "Alves (2024)", "30 ADHD, 30 controls", "Not fully extracted", "Not fully extracted", "Not fully extracted", "fMRI", "Analysis used blood oxygenation level-dependent (BOLD) time series data",
  "Amen (2021)", "1,006 ADHD, 129 controls", "Not fully extracted", "Not fully extracted", "Combined, Inattentive, and Hyperactive", "SPECT", "Large retrospective analysis; ADHD patients with no comorbidities; Controls had no psychiatric diagnosis, brain injury, or substance use",
  "Chaim-Avancini (2017)", "67 ADHD, 58 controls", "27.0 ± 6.0 (ADHD)", "52 males/15 females (ADHD)", "36 Inattentive, 31 Combined", "MRI", "Education: 13.5 ± 2.6 years; Male-only subgroup (n=52) also analyzed",
  "Schneider (2014)", "N=427, 40% ADHD, 60% Other Psych", "40.9 years ± 15.7", "Not fully extracted", "DSM-IV criteria", "DTI", "Pediatric population (younger than 18 years); Focus on white matter tracts",
  "Wang (2023)", "50 ADHD, 50 controls", "Not fully extracted", "Not fully extracted", "Inattentive, Hyperactive/Impulsive, Combined", "fMRI", "Used ADHD-200 dataset; analysis using convolutional neural networks",
  "Wang (2013)", "23 ADHD, 23 controls", "35.14 ± 9.75 (ADHD), 32.04 ± 9.23 (Controls)", "18 males/5 females in both groups", "All Combined Type", "MRI", "Focus on resting-state fMRI; All patients met lifetime criteria for Combined Type ADHD",
  "Wolfers (2016)", "184 ADHD, 103 unaffected siblings, 128 controls", "Not fully extracted", "Not fully extracted", "Not fully extracted", "fMRI", "Used Stop-Signal Task paradigm to assess response inhibition; included unaffected siblings as comparison group",
  "Yao (2018)", "112 ADHD, 77 controls (adults); 34 ADHD, 28 controls (children)", "Adults: 25.93±4.86 (ADHD), 26.04±3.94 (controls); Children: 9.79±1.86 (ADHD), 10.29±1.67 (controls)", "Adults: 37F/75M (ADHD), 34F/43M (controls); Children: All male", "Not fully extracted", "fMRI", "Used two separate datasets (adults and children); Focus on default mode network"
)

# Use knitr::kable for better cross-format compatibility
knitr::kable(
  demographics_data_neuroimaging,
  col.names = c("Study", "Sample Size", "Age (Mean ± SD)", "Gender", "ADHD Subtypes", "Imaging Modality", "Key Demographic Notes"),
  caption = "Neuroimaging Studies Demographics Summary"
)
```

## Neuropsychological and Cognitive Tests

```{r}
```

### Overview of Current Research Findings

The research available in your collection provides substantial evidence regarding the effectiveness and limitations of neuropsychological tests for diagnosing ADHD in adults.
The evidence points to several consistent findings:

### Limited Standalone Diagnostic Value

Most studies indicate that neuropsychological tests alone have poor discriminative validity when used as standalone diagnostic tools for adult ADHD.
Pettersson et al. (2018) found that "neuropsychological tests have a poor ability to discriminate between patients diagnosed with ADHD and patients not diagnosed with ADHD." Similarly, research by Nikolas et al. (2019) found that "single test measures provided performed poorly in identifying ADHD participants."

### Better Results in Combination with Other Measures

The research consistently shows that neuropsychological tests provide the most value when used as part of a comprehensive assessment approach.
Nikolas et al. (2019) reported that "a combined approach using self and informant symptom ratings, positive family history of ADHD, and a reaction time (RT) variability measure correctly classified 87% of cases." This suggests that while neuropsychological tests have limitations as standalone measures, they can contribute meaningfully to diagnostic accuracy when used as part of a more comprehensive evaluation protocol.

### Specific Findings on Test Performance

#### Most Promising Measures

Several specific neuropsychological domains have shown more utility than others:

1.  **Reaction Time Variability**: Multiple studies identified RT variability as one of the most sensitive measures.
    Nikolas et al. (2019) found that "measures of working memory, sustained attention, response speed, and variability best discriminated ADHD and non-ADHD participants."

2.  **Sustained Attention Measures**: Tests that assess sustained attention, particularly computerized continuous performance tests (CPTs), showed better discrimination capabilities than other measures.

3.  **Working Memory Tasks**: Measures of working memory were consistently identified as having some utility in distinguishing adults with ADHD.

#### QbTest Performance

The QbTest, a computerized test designed to measure all three core ADHD symptoms (inattention, hyperactivity, and impulsivity), showed mixed results:

1.  **Brunkhorst-Kanaan (2020)** found that the QbTest has "low discriminative power" with an AUC (Area Under the Curve) of only 0.65, sensitivity of 76%, and specificity of 40%.
    They concluded that "the QbTest is not able to discriminate between ADHD patients and non-ADHD patients in an outpatient clinic."

2.  **Edebol et al. (2013)** reported more positive findings, with sensitivity of 86% and specificity of 83%.
    However, they noted that sensitivity dropped substantially when trying to differentiate between individuals with ADHD and other clinical groups.

3.  **Adamou et al. (2022)** advised "caution when interpreting QbTest+ results in clinical populations," as their study found that QbTest+ scores failed to differentiate between patients diagnosed with ADHD and those without the diagnosis.

4.  **Söderström et al. (2014)** found that "the self-rating scales exhibited high sensitivity values but very low specificity values" while "the QBImpulsivity and QBInattention variables \[showed\] high specificity values and low sensitivity values."

### Challenges and Limitations

#### 1. Overlap with Other Clinical Conditions

A significant challenge in using neuropsychological tests for ADHD diagnosis is the overlap of cognitive deficits with other psychiatric conditions:

-   Pettersson et al. (2018) noted that "when adults with ADHD are compared with other psychiatric patients, the results are more inconsistent."
-   Nikolas et al. (2019) highlighted the "substantial overlap between ADHD symptoms and cognitive symptoms of other mental health conditions, such as depression and anxiety."

#### 2. Heterogeneity of ADHD Presentation

The neuropsychological profile of adults with ADHD is heterogeneous:

-   Nikolas et al. (2019) found that "adults with above average global cognitive abilities may underperform on some of these tasks relative to their IQ, even though their performance may be within an average range when compared with a broader population of adults."
-   This suggests that using standard cutoff scores may not be appropriate, particularly for high-functioning adults or those with higher education levels.

#### 3. Test Sensitivity and Ecological Validity

-   Sensitivity and specificity issues persist across most tests, with no single measure providing both high sensitivity and specificity.
-   Ecological validity remains a concern, as performance in structured testing environments may not reflect real-world functioning.

### Optimal Assessment Approach

Based on the research findings, the optimal approach for diagnosing adult ADHD appears to involve:

1.  **Comprehensive Clinical Interview**: All studies emphasized the importance of a thorough clinical interview as the foundation of assessment.

2.  **Structured Rating Scales**: Both self-report and informant-report scales provide valuable information about symptom presentation.

3.  **Selective Neuropsychological Testing**: Rather than extensive batteries, focused testing of sustained attention, working memory, and reaction time variability appears most useful.

4.  **Performance Validity Testing**: Including performance validity measures is essential, particularly given concerns about symptom exaggeration in adult ADHD assessment.

5.  **Family History Assessment**: Several studies found that including family history information significantly improved diagnostic accuracy.

### Conclusion

The research suggests that neuropsychological tests have a limited but valuable role in adult ADHD assessment.
While they should not be relied upon as standalone diagnostic tools, selective neuropsychological tests focusing on sustained attention, reaction time variability, and working memory can contribute to diagnostic accuracy when used as part of a comprehensive assessment approach that includes clinical interviews, rating scales, and family history.

The QbTest specifically appears to have some utility in quantifying ADHD symptoms, but its diagnostic accuracy is insufficient to serve as a primary diagnostic tool, particularly when attempting to differentiate ADHD from other psychiatric conditions.
The test may be most useful for tracking treatment response or quantifying symptom severity rather than establishing diagnosis.

These findings align with the broader consensus in the field that adult ADHD diagnosis should involve multiple assessment methods, with a structured clinical interview remaining the cornerstone of the diagnostic process.

```{r}
# Load the data but handle it as a simple table, not using gt for compatibility
df_neuro <- vroom::vroom("neuro.csv")

# Create a simple knitr::kable table that works in all output formats
knitr::kable(
  df_neuro,
  caption = "Demographics Table for Adult ADHD Review: Neuropsychological Tests",
  format = "pandoc"
)
```

The document mentions several other assessment tools that are commonly used in adult ADHD diagnosis, though without specific sensitivity/specificity data:

1.  **Structured Interviews:**
    -   Diagnostic Interview for ADHD in Adults (DIVA 2.0)
    -   Schedule for Affective Disorders and Schizophrenia for School-Age Children-Present and Lifetime Version (K-SADS-PL) adapted for adults
    -   Wender Reimherr Interview (WRI)
2.  **Self-Report Rating Scales:**
    -   Adult ADHD Self-Report Scale (ASRS-V1.1)
    -   Conners' Adult ADHD Rating Scales (CAARS) - both self and observer versions
    -   Wender Utah Rating Scale (WURS)
    -   Brown Attention-Deficit Disorder Scales
3.  **Additional Neuropsychological Tests:**
    -   Wisconsin Card Sorting Test
    -   Digit Span (WAIS)
    -   Sustained Attention to Response Task (SART)
    -   D-KEFS (Delis-Kaplan Executive Function System)

### Research Findings on Assessment Approaches

The document emphasizes several key points about neuropsychological assessment of adult ADHD:

1.  **Single tests have limited utility:** Individual neuropsychological measures typically show moderate to low sensitivity (19-58%) with higher specificity (72-100%), making them better at ruling in rather than ruling out ADHD.

2.  **Battery approaches are superior:** Using multiple tests or composite measures substantially improves diagnostic accuracy, with sensitivity and specificity both potentially reaching 90%.

3.  **Combined methods are most effective:** Integrating subjective measures (rating scales, interviews) with objective neuropsychological tests provides the highest diagnostic accuracy.

4.  **Reaction time variability (RTV)** emerges as one of the most consistent neuropsychological markers in adult ADHD across studies.

5.  **Comorbidities reduce diagnostic accuracy:** The presence of other conditions like bipolar disorder or borderline personality disorder significantly decreases the sensitivity of neuropsychological measures.
